---
title: "College Retention Across Sectors"
author: "Maritza Ruiz and Steven Todd McIntyre II"
abstract: "While enrollment in higher education is decreasing, it is important to understand the relationships that different aspects of higher education institutions and enrollment patterns have with student retention. In this paper, we examine the relationships between institution type, starting intensity, and second fall retention rates in the United States from 2015-2023. We find that public and private 4-year institutions consistently have the highest retention rates, while private non-profit institutions consistently have the lowest retention rates. Furthermore, students starting at full time status are more likely in all sectors to be retained into the second fall. These findings can aide higher education policymakers in understanding both what can lead to student success and to evaluate part-time and non-credential status programs for possible improvements in retention."
format: 
  html:
    toc: true
    code-fold: true
bibliography: references.bib
csl: https://www.zotero.org/styles/apa
execute:
  warning: false
---
# Research Questions
What is the relationship between institutional type (public vs. private, 2-year vs. 4-year) and student retention?

How does enrollment intensity (full-time vs. part-time) interact with institutional type to influence persistence and retention?

# Introduction
At a time when enrollment in higher education is declining, understanding how starting enrollment intensity (full-time vs. part-time) influences persistence and retention helps institutions identify barriers that affect student success and degree completion. 

Colleges and universities serve increasingly diverse student populations, many of whom attend part-time due to financial, work, or family obligations. Examining differences across institutional types can reveal how structural and resource differences within higher education shape student outcomes.Research has consistently shown that first-generation and low-income students face higher risks of attrition, with dropout rates varying across institutional contexts [@CataldiEtAl2018]. For example, national data indicate that first-generation students are more likely to leave college early, particularly at public and two-year institutions where resources are more limited [@Riehl1994]. 

Recent statistics further highlight that nearly 40% of college dropouts are first-generation students, underscoring the importance of institutional type in shaping persistence [@Bouchrika2024]. State-level data also demonstrates how retention rates differ across public institutions, with Oregon’s Higher Education Coordinating Commission reporting that 83% of students at four-year public institutions continued after their first year [@HECC2018]. 

These findings suggest that institutional type and enrollment intensity interact to influence student success, making it critical for higher education leaders to design targeted retention strategies, improve academic support systems, and create flexible pathways that promote persistence for all students, regardless of their enrollment status.

# Methods

This analysis utilized R [@R] and the following R libraries: tidyverse [@tidyverse], rio [@rio], knitr [@knitr], here [@here], and janitor [@janitor].
```{r}
# load libraries!
library(tidyverse)
library(rio)
library(here)
library(knitr)
library(janitor)
library(ggplot2)

# must use:
# [x]   pivot_longer
# [x]    mutate
# [x]    select
# [x]    filter
# [x]    pivot_wider
# [x]    group_by
# [x]    summarize

# import and clean data
overall_persistence <- import(
    here("data/PersistenceandRetention2025_Appendix.xlsx"), 
    sheet = "1. Overall", 
    skip = 2
  ) |>
  pivot_longer(
    cols = starts_with("20"),
    names_to = "entering_cohort_year",
    values_to = "rate"
  ) |>
  pivot_wider(
    names_from = "Metric",
    values_from = "rate"
  ) |>
  clean_names() |>
  mutate(entering_cohort_year = as.numeric(entering_cohort_year))

#  overall_persistence TEST CODE
  
#  names(overall_persistence) TEST CODE

# Select()columns for analysis
  overall_persistence_selected <- overall_persistence |>
    filter(starting_enrollment_intensity == "All") |>
    dplyr::select(entering_cohort_year, sector, second_fall_retention)
  
# Groupby() summary of average retention by   #sector across all years
sector_summary <-   overall_persistence_selected |>
  group_by(sector) |>
  summarize(mean_retention = mean(second_fall_retention, na.rm = TRUE))

```
We use data provided by the National Student Clearinghouse [@NationalStudentClearinghouse] to examine and provide descriptive statistics on second fall retention trends from 2015-2023 across US higher education institutions. We were particularly interested in grouping the analysis by institution sector (private for-profit 4-year, private nonprofit 4-year, public 2-year, public 4-year, and public PAB) and the intensity students started at (full time, part time, non-credential).

The National Student Clearinghouse Research Center’s retention data comes from enrollment records submitted directly by colleges and universities across the U.S. It tracks whether first-time students remain enrolled at their original institution (retention) or at any institution (persistence) into their second year. Data is collected from participating institutions’ student information systems and covers all sectors and enrollment intensities.

# Results

```{r}
#Bar chart: mean retention by sector
#Need to make title larger
sector_summary |>
  kable(
    caption = "Average Second Fall Retention by Sector (2015-2023)", 
    col.names = c("Sector", "Mean Retention"),
    digits = 2
  )

pub_4_mean_ret <- sector_summary$mean_retention[sector_summary$sector == "Public 4-year"]

pri_np_4_mean_ret <- sector_summary$mean_retention[sector_summary$sector == "Private nonprofit 4-year"]

pub_2_mean_ret <- sector_summary$mean_retention[sector_summary$sector == "Public 2-year"]

pub_pab_mean_ret <- sector_summary$mean_retention[sector_summary$sector == "Public PAB"]

pri_fp_4_mean_ret <- sector_summary$mean_retention[sector_summary$sector == "Private for-profit 4-year"]
```

Retention rates vary significantly by sector. Among individual sectors, "Public 4-year" institutions and "Private nonprofit 4-year" institutions demonstrated higher mean retention rates from 2015-2023 (`r round(pub_4_mean_ret, digits = 2)` and `r round(pri_np_4_mean_ret, digits = 2)`, respectively), suggesting these types of institutions may offer more consistent support structures or resources that promote student persistence.
In contrast, "Private for-profit 4-year" institutions have the lowest mean retention rate over this period, at `r round(pri_fp_4_mean_ret, digits = 2)`, which may reflect challenges such as limited student services, financial instability, or differences in student demographics. "Public 2-year" and "Public PAB" sectors, at `r round(pub_2_mean_ret, digits = 2)` and `r round(pub_pab_mean_ret, digits = 2)`, respectively, fell just above "Private for-profit 4-year", showing moderate retention outcomes.
This distribution highlights the importance of sector-specific strategies when addressing student retention. Institutions with lower rates may benefit from targeted interventions, while those with higher rates could serve as models for effective retention practices.

```{r}
# create visualization A: 
overall_persistence |>
  filter(starting_enrollment_intensity == "All") |>
  ggplot(aes(x=entering_cohort_year, y=second_fall_retention)) +
  geom_line(aes(color = sector)) +
  labs(
    title = "Second Fall Retention by Sector",
    x = "Cohort Year",
    y = "Retention Rate",
    color = "Sector"
  ) +
  scale_color_discrete(breaks= c('Public 4-year', 'Private nonprofit 4-year', 'All', 'Public PAB', 'Public 2-year', 'Private for-profit 4-year')) +
  theme_minimal() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.line = element_line(color = "black"),
    axis.ticks = element_line(color = "black"),
    plot.title = element_text(size = 11, hjust = 0.5)
  )

```

Retention rates show differences in the various sectors over time. Again, private nonprofit and public 4-year institutions consistently maintain higher second-fall retention compared to other institutions, with private for-profit and public 2-year institutions showing the lowest overall retention rates. That said, for-profit institutions have seen an increase in retention from 2020 to 2023, even surpassing public 4-year institutions in 2023.

```{r}
overall_persistence |>
  filter(starting_enrollment_intensity != "All") |>
  group_by("Starting Intensity" = starting_enrollment_intensity, "Sector" = sector) |>
  summarize("Average Second Fall Retention (2015-2023)" = mean(second_fall_retention)) |>
  ggplot(aes(x = `Sector`, y = `Average Second Fall Retention (2015-2023)`, fill = `Starting Intensity`)) +
  geom_col(position = position_dodge()) +
  labs(
    title = "Average Second Fall Retention by Sector and Starting Intensity (2015-2023)",
    x = "Sector",
    y = "Mean Retention Rate",
    fill = "Starting Intensity"
  ) +
  theme_minimal() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.line = element_line(color = "black"),
    axis.ticks = element_line(color = "black"),
    plot.title = element_text(size = 8),
    axis.text.x = element_text(size = 6, angle = 45, hjust = 1)
  )
```
Across sectors, it looks like full-time students are substantially more likely to be retained into the second fall. Part-time status seems to have a relatively negligible effect on retention as compared to non-credential status, which may call into question the importance of degree-seeking status for non-full time students.

# Discussion
The findings of this highlight the influence of both institutional type and enrollment intensity on student retention across U.S. higher education sectors. Consistent with prior research, public and private four-year institutions demonstrate the highest second-fall retention rates, suggesting that these institutions may provide more robust academic support systems, financial aid opportunities, and campus engagement structures that foster persistence. In contrast, private for-profit and public two-year institutions continue to face challenges in retaining students, reflecting structural differences in mission, funding, and student demographics.

The increase in retention rates among for-profit institutions between 2020 and 2023 is particularly noteworthy. While these institutions have historically been reviewed for low completion rates and high student debt burdens, the upward trend suggests that targeted reforms or shifts in student support strategies may be yielding positive outcomes. This improvement warrants further investigation into the specific practices or policy changes that contributed to these gains, as they may offer transferable insight for other sectors struggling with retention.

Enrollment intensity remains a critical factor in student persistence. Full-time students consistently perform at higher retention rates across all sectors, reinforcing the importance of enrollment status as a predictor of success. However, the relatively weak retention outcomes among non-credential students highlight a pressing concern: students who enter higher education without a clear degree-seeking pathway may lack the institutional support necessary to persist. This finding raises important questions about how colleges and universities design programs for part-time and non-credential learners, particularly given the growing diversity of student populations balancing work, family, and financial obligations.

Together, these results suggest that institutional type and enrollment intensity interact in ways that shape student success. Public two-year institutions, for example, serve a disproportionate number of first-generation and low-income students, populations already at higher risk of attrition. Without adequate advising, financial aid, and flexible pathways, these students may struggle to persist, contributing to lower overall retention rates. Conversely, the stronger outcomes observed at four-year institutions may reflect both greater resource availability and more structured pathways to degree completion.

From a policy perspective, these findings highlight the need for sector-specific retention strategies. Institutions with lower retention rates—particularly public two-year and private for-profit colleges—may benefit from targeted interventions such as enhanced advising, expanded financial aid, and improved transfer pathways. At the same time, policymakers should consider how to support part-time and non-credential students, whose persistence is often undermined by structural barriers. Flexible scheduling, competency-based education, and integrated student services may help address these challenges.

Finally, the results contribute to the broader conversation about declining enrollment in higher education. Ensuring that students not only enter but remain and succeed within higher education is essential to reversing enrollment declines. By identifying where retention gaps are most pronounced, institutions and policymakers can design interventions that promote equity, strengthen student success, and sustain the vitality of higher education in the United States.
